### 2017-03-31 DLM/writeup.tex
writeup changes

\begin{equation}
\label{Reparam}
\boldsymbol{\theta} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}.
\end{equation}

@@

\begin{equation}
\label{ReparamDist}
\log(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)})) = \log(|J|) + \log(q(\boldsymbol{\epsilon}))
\end{equation}

where $J$ is the Jacobian Matrix of the transformation $f$, a function of $\boldsymbol{\lambda}$ and $\boldsymbol{\epsilon}$. The derivative of (\ref{ELBO}) can be estimated by

\begin{equation}
\label{ReparamDeriv}
\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} \ln (p(y_{1:T}, f(\boldsymbol{\lambda}, \boldsymbol{\epsilon}_s))) - \nabla_{\boldsymbol{\lambda}}\log(|J(\boldsymbol{\lambda}, \boldsymbol{\epsilon}_s)|),
\end{equation}
where simulations of $\boldsymbol{\theta}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$. The variance of this estimator is often orders of magnitude lower than in (\ref{ScoreDeriv}) and in many cases $S=1$ is sufficient.

@@

\begin{align}
y_t &= \mu + x_t + \sigma_y \epsilon_t \label{DLM:Measure}\\
x_t &= \phi x_{t-1} + \sigma_x \nu_t \label{DLM:Transition}
\end{align}

@@

\begin{align}
\mu &\sim \mathcal{N}(\bar{\mu}, \tau) \\
\phi &\sim U(-1, 1) \\
\sigma^2_y &\sim \mathcal{IG}(\mbox{shape = }\alpha_y, \mbox{scale = }\beta_y) \\
\sigma^2_x &\sim \mathcal{IG}(\mbox{shape = }\alpha_x, \mbox{scale = }\beta_x) \\
x_0 &\sim \mathcal{N}(0, \sigma^2_x(1 - \phi^2)^{-1})
\end{align}

@@

We transform $\sigma^2_y$ and $\sigma^2_x$ so that $\boldsymbol{\theta} = \{\mu, \phi, \log(\sigma^2_y), \log(\sigma^2_x) \} \in \mathbb{R}^4$ and consider four different functional forms for the approximating distribution $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda})$:
\begin{enumerate}
\item $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda}) = \mathcal{N}(\boldsymbol{\psi}, \textbf{V})$, where $\textbf{V}$ is a diagonal matrix.
\item $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda}) = \mathcal{N}(\boldsymbol{\psi}, \boldsymbol{\Sigma})$, where $\boldsymbol{\Sigma}$ allows non-zero off-diagonal terms.
\item $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda})$ is an Inverse Autoregressive flow (Kingma, Salimans, Jozefowicz, Chen, Sutskever and Welling, 2016).
\item  $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda}) = \prod \left(q(\theta_i) q(x_{it})\right) c(\boldsymbol{\theta}, \textbf{x})$ where the form of the copula $c$ and marginals are informed by an out of date MCMC sample.
\end{enumerate}
The first three functional forms are transformation of standard normal noise variables while the fourth is a transformation of uniform noise variables.

@@

>>>>

\begin{equation}
\label{Reparam}
\boldsymbol{\theta} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}).
\end{equation}

@@

\begin{equation}
\label{ReparamDist}
\log(q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = \log(|J|) + \log(q(\boldsymbol{\epsilon}))
\end{equation}

where $J$ is the Jacobian Matrix of the transformation $f$, which may depend on $\boldsymbol{\lambda}$ and $\boldsymbol{\epsilon}$. The derivative of (\ref{ELBO}) can be estimated by

\begin{equation}
\label{ReparamDeriv}
\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} \ln (p(y_{1:T}, f(\boldsymbol{\lambda}, \boldsymbol{\epsilon}_s))) - \nabla_{\boldsymbol{\lambda}}\log(|J(\boldsymbol{\lambda}, \boldsymbol{\epsilon}_s)|),
\end{equation}
where simulations of $\boldsymbol{\theta}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$. The variance of this estimator is often orders of magnitude lower than the estimator in (\ref{ScoreDeriv}), often $S=1$ is sufficient.

@@

\begin{align}
y_t &= \gamma + x_t + \sigma_y \epsilon_t \label{DLM:Measure}\\
x_t &= \phi x_{t-1} + \sigma_x \nu_t \label{DLM:Transition}
\end{align}

@@

\begin{align}
\gamma &\sim \mathcal{N}(\bar{\gamma}, \sigma^2_{\gamma}) \\
\phi &\sim \mathcal{U}(-1, 1) \\
\sigma^2_y &\sim \mathcal{IG}(\mbox{shape = }\alpha_y, \mbox{scale = }\beta_y) \\
\sigma^2_x &\sim \mathcal{IG}(\mbox{shape = }\alpha_x, \mbox{scale = }\beta_x) \\
x_0 &\sim \mathcal{N}(0, \sigma^2_x(1 - \phi^2)^{-1})
\end{align}

@@

We transform $\sigma^2_y$ and $\sigma^2_x$ so that $\boldsymbol{\theta} = \{\gamma, \phi, \log(\sigma^2_y), \log(\sigma^2_x) \} \in \mathbb{R}^4$ and consider four different functional forms for the approximating distribution $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda})$:
\begin{enumerate}
\item $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda}) = \mathcal{N}(\boldsymbol{\psi}, \textbf{V})$, where $\textbf{V}$ is a diagonal matrix.
\item $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda}) = \mathcal{N}(\boldsymbol{\psi}, \boldsymbol{\Sigma})$, where $\boldsymbol{\Sigma}$ allows non-zero off-diagonal terms.
\item $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda})$ is an Inverse Autoregressive flow (Kingma, Salimans, Jozefowicz, Chen, Sutskever and Welling, 2016).
\item  $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda}) = \prod \left(q(\theta_i) q(x_{it})\right) c(\boldsymbol{\theta}, \textbf{x})$ where the form of the copula $c$ and marginals are informed by an out of date MCMC sample.
\end{enumerate}
The first three functional forms are transformation of standard normal noise variables while the fourth is a transformation of uniform noise variables.

@@

\section{Appendix}

\subsection{Variational Approximations for the Dynamic Linear Model}

The joint distribution $p(y_{1:T}, \boldsymbol{\theta}, x_{1:T})$ for the Dynamic Linear Model is given by

\begin{align}
p(y_{1:T}, \boldsymbol{\theta}, x_{1:T}) &= \prod_{t=1}^{T} \left( \mathcal{N}(y_t | x_t + \gamma, \sigma^2_y) \mathcal{N}(x_t | \phi x_{t-1}, \sigma^2_x) \right) \mathcal{N}(x_0 | 0, \sigma^2_x (1 - \phi^2)^{-1}) \nonumber \\
&\mathcal{N}(\gamma | \bar{\gamma}, \sigma^2_{\gamma}) \mathcal{U}(\phi | -1, 1) \mathcal{IG}(\sigma^2_y | \alpha_y, \beta_y)\mathcal{IG}(\sigma^2_x | \alpha_x, \beta_x) 
\end{align}
with 
\begin{align}
\log(p(y_{1:T}, \boldsymbol{\theta}, x_{1:T})) &= -(T/2 + \alpha_y + 1) \log(\sigma_y^2) -(T/2 + \alpha_x + 3/2) \log(\sigma_x^2) \nonumber \\
&- \frac{\sum_{t=1}^{T}(y_t - \gamma - x_t)^2}{2 \sigma^2_y} - \frac{x_0^2(1 -\phi^2) + \sum_{t=1}^{T}(x_t - \phi x_{t-1})^2}{2 \sigma^2_x} \nonumber \\
&- \frac{(\gamma - \bar{\gamma})^2}{2 \sigma^2_{\gamma}} - \frac{\beta_y}{\sigma^2_y} - \frac{\beta_x}{\sigma^2_x} + c \label{logjoint}. 
\end{align}

We use Kingma and Welling's (2014) reparameterised approach to the variational approximation in all of our applications.

\subsubsection{The Multivariate Normal Approximations}

The distribution $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ for the first two approaches is defined by the following transformation, where $\epsilon_{1}, \dots, \epsilon_{T+5}$ have independent standard normal distributions.
\begin{align}
\sigma^2_y &= \exp(\mu_1 + L_{11} \epsilon_1)  \label{transform1} \\
\sigma^2_x &= \exp(\mu_2 + L_{21} \epsilon_1 + L_{22} \epsilon_2) \label{transform2} \\
\phi &= \mu_3 + L_{31} \epsilon_1 + L_{32} \epsilon_2 + L_{33} \epsilon_3 \label{transform3} \\
\mu &= \mu_4 + L_{41} \epsilon_1 + L_{42} \epsilon_2 + L_{43} \epsilon_3 + L_{44} \epsilon_4 \label{transform4} \\
x_t &= \mu_{t+5} + \sum_{i=1}^{t+5} L_{t+5, i} \epsilon_i \mbox{ for } t = 0, \dots, T \label{transform5} 
\end{align}

The approximation is parameterised by $\boldsymbol{\lambda} = \{ \boldsymbol{\mu}, \textbf{L} \}$ where $\textbf{L}$ is the lower triangular decomposition of the Multivariate Normal variance matrices $\textbf{V}$ and $\boldsymbol{\Sigma}$. In the case where the diagonal matrix $\textbf{V}$ is used, restrict $L_{ij} = 0$ for $i \neq j$.
The log determinant of the Jacobian is given by
\begin{equation}
\label{logdetjac}
\log(|J|) = - \left(\sum_{i = 1}^{i = T+5} \log(L_{ii}) + \mu_1 + L_{11} \epsilon_1 + \mu_2 + L_{21} \epsilon_1 + L_{22} \epsilon_2\right).  
\end{equation}

Derivatives of (\ref{logjoint}) with respect to $\boldsymbol{\theta}$ and $x_{0:T}$, as well as derivatives of (\ref{transform1}) - (\ref{logdetjac}) with respect to $\boldsymbol{\mu}$ and $\textbf{L}$ are easily obtained for use in estimating the gradient of the ELBO.
